Synthesizing Quotient Lenses

### 2017-10-11 papers/QuotientLensSynth/main.tex
Made more changes

However, quotient lenses are sometimes difficult to program because the
equivalence relations defined on the data, as well as the underlying lens which
operates on the equivalence classes are complicated. These two challenges make
quotient languages impractical to use in many cases spite of their advantages.

In this paper, we address both of these challenges. First, we propose the
language of {\em Quotient Regular Expressions} or QREs, which enables the
programmer to explicitly define a broad class of equivalence relations on
regular languages using a compact and convenient notation. We then define a
class of {\em QRE lenses}, a class of quotient lenses which are typed using
QREs, and show how to {\em synthesize} QRE lenses from their QRE types and a
set of input-output examples. Finally, we demonstrate the applicability of this
approach by synthesizing QRE lenses between several real-world data formats
from the {\tt data.gov} database.
\end{abstract}

@@

in the EndNote format, and vice versa.

@@

However, the constraints that are imposed by the lens laws are sometimes too
strict. For instance, the programmer may wish to ignore updates to the data
that only change the order of fields, or add whitespace characters that in reality are not
part of the data. Foster et al therefore proposed \textit{quotient lenses}
~\cite{quotient} as a solution to this problem. Quotient lenses enable a
programmer to write lenses which transform data that is quotiented out by an
equivalence relation, so that records which are equivalent to each other modulo
this relation are considered the same by the underlying lens.

Unfortunately, quotient lenses are sometimes difficult to program because the
equivalence relations defined on the data, as well as the underlying lens which
operates on the equivalence classes are complicated. For example, suppose that
the programmer wants to program the \bibtex to EndNote transformation in such a
way that the program allows for any of two representations of the authors
name (e.g. ``Conway, J. H." and ``Conway J. H"), ignores the order in which the
author and title fields occur in each record, and allows for arbitrary amounts
of whitespace in between the author and title fields. These rules allow for
these two \bibtex citation records to be considered the same:

@@

The Boomerang quotient lens which does the \bibtex to EndNote conversion is

\sam{TODO: PUT BOOMERANG EXPRESSION HERE!},

an expression which is difficult to write and understand.

We address the challenge of programming quotient lenses in two ways. First, we
propose the language of {\em Quotient Regular Expressions} or QREs, which
enables the programmer to explicitly define a broad class of equivalence
relations on regular languages using a compact and convenient notation. For
example, the QRE which describes the equivalence relation on the \bibtex
citation records is as follows:

\sam{TODO: PUT QRE FOR BIBTEX HERE!}

This expression better describes the equivalence relation since it is explicit
and more precise.

However, even though QREs enable the programmer to describe complicated
equivalence relations on data more simply, the underlying lens which maps the
equivalence classes of the \bibtex records to the equivalence classes of the
EndNote records is also challenging to program, mostly because of the verbosity
of the expression:

\sam{TODO: PUT QRE LENS FOR BIBTEX HERE!}

@@

Our approach is to define a class of {\em QRE lenses} which are the same as
bijective lenses with canonizers at the ends. That is, a QRE lens uses the
source QRE to compute a unique representative for a datum modulo the
equivalence relation defined by the QRE before applying a bijective lens to
this representative in the forward direction, and operates similarly same in the
backward direction but using the view QRE.

@@

$$\mathit{synth} \; \mathtt{BIBTEX} \; \mathtt{ENDNOTE}$$

to derive a quotient lens that does the \bibtex to EndNote.

@@

\sam{Should we have a special construct like squash-synth(R, R') in addition to
squash(R, R', l) or maybe even just have only squash-synth(R, R')? Perhaps
having to explicitly write the lens weakens our case that QREs are compact and
convenient}

>>>>

However, quotient lenses are sometimes difficult to program, firstly because the
programmer has to give a specification for the data as well as give a separate
specification for the equivalence relations defined on the data. Secondly, even
after giving both of these specifications, the programmer also has to manually
write the underlying lens which maps between the respective equivalence
classes, a task which is itself challenging.

In this paper, we address both of these challenges. First, we propose the
language of {\em Quotient Regular Expressions} or QREs, which enables the
programmer to explicitly define a broad class of equivalence relations
directly on regular languages using a compact and convenient notation, and hence
negates the need to specify the data and the equivalence relation on the data
separately. We then define a class of {\em QRE lenses}, a class of quotient
lenses which are typed using QREs, and show how to {\em synthesize} QRE lenses
from their QRE types and a set of input-output examples. Finally, we
demonstrate the applicability of this approach by synthesizing QRE lenses
between several real-world data formats from the {\tt data.gov} database.
\end{abstract}

@@

\noindent in the EndNote format, and vice versa.

@@

However, the constraints that are imposed by the lens laws are sometimes too
strict. For instance, the programmer may wish to ignore updates to the data
that only change the order of fields, or add whitespace characters that in
reality are not part of the data. Foster et al therefore proposed
\textit{quotient lenses} ~\cite{quotient} as a solution to this problem
Quotient lenses, which are implemented as a refinement of the bidirectional
string processing language Boomerang, enable a programmer to write lenses which
transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

\sam{Go over quotient lens laws?}

Unfortunately, quotient lenses are sometimes difficult to program in Boomerang,
firstly because the programmer has to give a specification for the data as well
as give a separate specification for the equivalence relations defined on the
data. For example, suppose that the programmer wants to program the \bibtex to
EndNote transformation in such a way that the program allows for any of two
representations of the authors name (e.g. ``Conway, J. H." and ``Conway J. H"),
ignores the order in which the author and title fields occur in each record,
and allows for arbitrary amounts of whitespace in between the author and title
fields. These rules allow for these two \bibtex citation records to be
considered the same,

@@

with the EndNote records having a similar equivalence relation defined on them.

In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map each element of a regular language to a
canonical representation of that element. Therefore, in order for the programmer
to give a specification for the \bibtex records, then the programmer
has to write a regular expression which matches all possible representations of
\bibtex records, and then program a set of canonizers which does the job of
quotienting the data. This often makes the code verbose and error-prone.

\sam{TODO: PUT BOOMERANG CANONIZERS HERE!}

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which enables the programmer to explicitly define a broad
class of equivalence relations directly on regular languages using a compact
and convenient notation, and hence negates the need to specify the data and the
equivalence relation on the data separately. For example, a QRE which
specifies the equivalence relation on the \bibtex citation records is as
follows:

\begin{verbatim}
let bib_author  = 
"author = {" . squash bib_space_names -> bib_comma_names using 
(get (synth bib_space_names <=> bib_comma_names)) . "}"

let bib_ignore_wsp =  project ("," . [ \n\t\r]* ) -> ",\n"

let bib_perm : canonizer = perm (label, bib_author, bib_title) with
bib_ignore_wsp

let bibtex  = "@book{" . bib_perm . "}"
\end{verbatim}

\sam{Should we have a special construct like squash-synth(R, R') in addition to
squash(R, R', l), or maybe even just have only squash-synth(R, R')?}

The first expression specifies the author field as
having one of two representations (\verb!bib_space_names! and
\verb!bib_comma_names!), with the \verb!bib_comma_names! being the
canonical representation and the expression

\noindent \verb!(get (synth bib_space_names <=> bib_comma_names))! giving the
function which converts names described by \verb!bib_space_names! to their canonical
form described  by \verb!bib_comma_names!.

The second expression specifies the regular language which is a comma followed
by arbitrary occurences of whitespace characters quotiented by the equivalence
relation which makes all members of this language equivalent, with the canonical
representation being a comma followed by a newline character.

The third expression specifies that the label, title and author fields
are can occur in any order, with each of these fields separated by a comma and
arbitrary amounts of whitespace. The canonical representaion is that in which
the label, author and title fields occur in that order, separated by a comma and
a newline.

However, even though QREs enable the programmer to specify the data as well as
the equivalence relation on the data in the same expression, the programmer
still has to manually program the underlying lens which maps the equivalence
classes of the \bibtex records to the equivalence classes of the EndNote
records

\sam{TODO: PUT BIJECTIVE LENS HERE!},

\noindent task which is itself challenging and erro-prone.

@@

Our approach is to define a class of {\em QRE lenses} which are the same as
bijective lenses with canonizers at the ends. That is, a QRE lens uses the
source QRE to compute a unique representative for the data modulo the
equivalence relation defined by the QRE before applying a bijective lens to
this representative in the forward direction, and operates similarly same in the
backward direction, but using the view QRE.

@@

$$\mathit{synth} \; \verb!bibtex! \; \verb!endnote!$$

to derive a quotient lens that does the \bibtex to EndNote transformation.

@@
